\emph{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings

### 2020-05-11 paper_elsevier/paper_elsevier.tex
edits before send to Jodi

  \journal{Dendrochronologia} % Sets Journal name

@@

\usepackage{amsmath}
\usepackage{float}
\usepackage{booktabs}

@@

  \begin{abstract}
  We present a new R package to provide dendroecologists with tools to infer, quantify, analyze, and visualize growth suppression events in tree rings caused by insect defoliation, and to reconstruct defoliator outbreak chronologies. \texttt{dfoliatR} is based on the OUTBREAK program and builds on existing resources in the \texttt{R} and \texttt{dplR} computing environments. It performs an indexing procedure to remove climatic signals, represented in non-host chronologies or other reconstructed climate series, from the host-tree series. It then infers defoliation events in individual trees based on user-specified thresholds. Site-level analyses identify outbreak events that synchronously affect user-defined numbers or proportions of involved host trees. Functions are provided for summary statistics and graphics of tree- and site-level series. \texttt{dfoliatR} expands on OUTBREAK to provide greater control of suppression thresholds, additional output tables, and high-quality graphics. We provide two example data sets and script to enable users to gain familiarity with the package and its capabilities. The source code is available in the Comprehensive R Archive Network (CRAN) and on GitHub.
  \hfill\break
  \end{abstract}
   \begin{keyword} Dendroecology, spruce budworm, \emph{Choristoneura}, pandora moth, \emph{Coloradia pandora Blake}, larch-bud-moth
\newpage\end{keyword}
 \end{frontmatter}

@@

Variation in the width and morphology of annual radial growth rings in trees permits dating and quantification of past forest insect defoliator outbreaks. Defoliation can be distinguished from climate- and other disturbance-related influences by comparing ring-width or other annually-resolved features in the wood of host species to that of non-host species or annually-resolved climate records. The effect of defoliation on radial growth of trees has been recognized since the 1860s, and used to reconstruct outbreak regimes since the 1950s (Blais 1954; Lynch 2012). It was not until the 1980s, however, that precise dendrochronological techniques were applied for inferring defoliation events and reconstructing defoliator outbreak regimes (Swetnam, Thompson, and Sutherland 1985; Speer 2010; Lynch 2012). The first studies (Swetnam, Thompson, and Sutherland 1985; Swetnam and Lynch 1989, 1993) focused on developing historical reconstructions of western spruce budworm (WSBW; \emph{Choristoneura freemani} (Lepidoptera: Tortricidae); previously known as \emph{C. occidentalis}) outbreak regimes. The methodology has since been successfully applied to a wide range of defoliator species, most of which are conifer herbivores, and has evolved in sophistication for a wide range of ecosystem situations (Lynch 2012).

The main dendrochronological tool for inferring, dating, and characterizing defoliator outbreaks from tree-ring records has been the software routine OUTBREAK (Swetnam, Thompson, and Sutherland 1985; Holmes and Swetnam 1986; Swetnam and Lynch 1989). OUTBREAK computes indices (described later in detail) of suppressed growth by subtracting a detrended and standardized climate series (a ``control'' chronology) from individual host-tree detrended and standardized radial growth series after the host and non-host series have been brought to a common variance. The non-host chronology usually consists of a site chronology developed from non-host tree species growing on a climate-sensitive site, but a gridded climate data point series, like the North American Drought Atlas (Cook and Krusic 2004) also suffices. If the host and non-host species respond similarly to climate (which can and should be tested), the derived series retains variability that the host and non-host series do not have in common, generally the insect signal and some unexplained variability (noise). The user defines a rule base specifying the magnitude and duration that a period of indexed growth suppression must meet or surpass for a period of suppressed growth to be inferred as a defoliation event at the tree level.

Though powerful, OUTBREAK is outdated and increasingly difficult to use in modern computing environments. It was written in FORTRAN V with inherently severe restrictions, as RAM and disk space were limited at that time (256 kb and 10 MB, respectively) and FORTRAN conventions imposed very strict formatting, file naming, and output conventions. The program lacks a graphical interface or capabilities, forcing users to import generated text files into spreadsheets or other software to assess results and performing analyses. Furthermore, OUTBREAK can only handle one test at a time, creating barriers to batch operation and a large burden for researchers with a datasets including multiple sites. We developed \texttt{dfoliatR} as an \texttt{R}- and \texttt{dplR}-based library to overcome these issues.

\texttt{dfoliatR} adds to a growing suite of dendrochronology packages in the \texttt{R} computing environment (R Core Team 2019). Stemming from the \texttt{dplR} library (Bunn 2008) that enables \texttt{R} users to read and write an array of tree-ring data formats, standardize ring width series, build and evaluate chronologies, and perform quality control (to name a few), one can now also measure ring widths from scanned images of prepared samples (Shi et al. 2019; Lara, Bravo, and Sierra 2015), conduct and check crossdating (Bunn 2010), analyze sub-annual anotomical features (Campelo et al. 2016; Rathgeber et al. 2011), and perform many analytical tests (Jevšenak and Levanič 2018; Zang and Biondi 2015). Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events (\texttt{TRADER}: Altman et al. 2014), metrics of growth resilience (\texttt{pointRes}: van der Maaten-Theunissen, van der Maaten, and Bouriaud 2015), and fire history (\texttt{burnr}: Malevich, Guiterman, and Margolis 2018). The key objective of \texttt{dfoliatR} is to provide tools to identify and analyze insect defoliation and outbreak events by building on the methods employed by OUTBREAK. It capitalizes on the robust software already available in \texttt{R} by using \texttt{dplR} data formats for incoming tree-ring series and providing output data formats embodied by the \texttt{tidyverse} (Wickham et al. 2019) that include efficient data manipulation (Wickham et al. 2020) and graphics (Wickham 2016).

In this paper, we describe the statistical methods employed by \texttt{dfoliatR}, compare results to those produced by OUTBREAK, and present an example analysis inlcuding test data sets and script. Users need not have much experience in \texttt{R} to replicate the analyses and graphics as presented. The \texttt{R} code below is executable in an \texttt{R} session once the required libraries are installed and loaded. Support documentation in addition to this paper is provided within the package via standard help menus and on the package website (\url{https://chguiterman.github.io/dfoliatR/}), which includes up-to-date vignettes that describe various routines. Code to generate a preprint of this manuscript, including the \texttt{R} scripts and tubular and graphical output is available from \url{https://github.com/chguiterman/dfoliatR_paper}.

@@

Users can develop these data sets in the software of their choosing, such as \texttt{dplR} or ARSTAN (Cook and Holmes 1996). It is important that the host-tree data include only one tree-ring series per tree. Both \texttt{dplR} and ARSTAN have options for averaging multiple sample series into a tree-level series.

@@

@@

Negative departures in the normalized GSI (NGSI; GSI converted to z-scores) that surpass user-specified thresholds in duration and magnitude are defined as \emph{defoliation events}. As in OUTBREAK, NGSI must reach the magnitude threshold, assessed for each year within the departure sequence. The default setting is -1.28 (NGSI is in units of standard deviation), which was previously determined to be representative of WSBW effects (Swetnam and Lynch, 1989) and is commonly used for other species. Failure to reach the specified threshold results in sequences not being inferred as defoliation events.

Event duration is assessed by examining sequences of negative NGSI (that exceed the magnitude threshold) before and after the year of maximum departure. Each defoliation event is allowed one single-year positive excursion on each side of the year of maximum departure. Duration is computed across the entire negative sequence that allows these two positive excursions. As in OUTBREAK, the user specifies a duration threshold (minimum number of years) for a departure sequence to be inferred as a defoliation event. The default threshold is eight years, as is commonly used in WSBW studies (Swetnam and Lynch, 1989). Failure to reach the specified duration threshold results in sequences not being inferred as defoliation events, regardless of NGSI values (i.e., both thresholds must be met). Researchers can, and should, adjust the duration and magnitude parameters accordingly and critically evaluate the results, as insect species vary in the length of their outbreaks and the degree to which they can suppress tree growth.

@@

Diverging from OUTBREAK, \texttt{dfoliatR} allows users to extend defoliation events on individual trees by bridging between sequential events. In cases where two defoliation events are separated by a single year, bridging will link them into a single event. This option was added to \texttt{dfoliatR} during the testing phase of development, when we realized that OUTBREAK deliberately omits sequential, or back-to-back events, even when these events surpass thresholds in duration and magnitude. Instead, OUTBREAK will select the one sequential event with the greatest negative departure year. In every case we assessed (described below) we felt that the OUTBREAK-omitted defoliation events should have been maintained and recorded. Due in large part to reconstructions using OUTBREAK (see papers cited by Lynch 2012), we now know considerably more about forest defoliator outbreak regimes than we did in the 1980s when OUTBREAK was written. We think that two or more prolonged events separated by a single year should in some situations be considered a single event. This is particularly relevant to WSBW and spruce budworm (\emph{C. fumiferana}), for which regimes have been reconstructed for several geographic areas and for which researchers and forest health experts have gained considerable knowledge (Schmitt, Grimble, and Searcy 1984; Sanders et al. 1985; Brookes et al. 1982, and many later publications). In many cases, the greatest growth suppression often occurs late in the outbreak due to cumulative effects accrued on a tree's resources (Swetnam and Lynch 1989, 1993). We urge caution in using the bridging option, however, because it may not be appropriate for all studied insects, such as in situations where impacted stands barely recover from one outbreak before another begins, as with pine processionary caterpillars (\emph{Thaumetopoea pityocampa} (Lepidoptera: Thaumetopoeidae)) (Carus 2004, 2009).

@@

Defoliation of one or a few trees does not constitute an outbreak. To determine when defoliation becomes an \emph{outbreak event}, \texttt{dfoliatR} composites the individual tree defoliation series into a site-level chronology with the \texttt{outbreak()} function. Users have options to define the number and/or the proportion of trees required for an event to be considered an outbreak. Three parameters control whether a defoliation event constitutes an outbreak: the minimum number of trees available, the minimum number of trees recording defoliation, and the percent of trees recording defoliation. The first allows the researcher to make a judgement call as to the confidence ascribed to reduced sample depth toward the ends of their chronologies, thus compensating for the ``fading record problem'' (Swetnam, Allen, and Betancourt 1999). The second two parameters adjust the scale of defoliation considered to be an outbreak. Absolute numbers of trees and percentages can be applied separately or in conjunction, following filtering conventions in tree-ring fire history studies (Malevich, Guiterman, and Margolis 2018). We urge users to carefully consider the choice of absolute numbers in situations where the number of trees represented in the series varies with time, or the choice of percentages when sample size is small.

@@

We tested \texttt{dfoliatR} against OUTBREAK by comparing NGSI to OUTBREAK's normalized corrected indices for individual trees and years, defoliation status for individual trees and years, and percentage of trees recording outbreaks at the site level, using standardized ring-width data from eight sites in British Columbia, Colorado, and New Mexico, and author-provided non-host site chronologies.

We detrended host data for both \texttt{dfoliatR} and OUTBREAK using ARSTAN (v6.1, Cook and Holmes 1996) with default double detrending (128 year wavelength and a 50\% smoothing spline). In both \texttt{dfoliatR} and OUTBREAK we used event thresholds of -1.28 normalized indices, 8 years duration, and allowed for events at the end of series in seven of eight sites. These seven sites were sampled during ongoing outbreak events (Swetnam and Lynch 1993). We found it necessary to be consistent in how we detrended and in what software (e.g., ARSTAN vs \texttt{dplR}) we employed because subtle differences in standardized ring-width indices generated between the programs transferred into differences between \texttt{dfoliatR} and OUTBREAK. In the end, we chose to only use the standardization output files from ARSTAN, which are easily read into \texttt{R} (and then \texttt{dfoliatR}) using the \texttt{dplR} package.

@@

\begin{itemize}
\item
  \emph{Series-end events} (40\% of the total) in which OUTBREAK included ``truncated outbreaks'' (for seven sites) at the end of each series. In \texttt{dfoliatR}, this option is controlled by the ``series\_end\_events'' parameter to \texttt{defoliate\_trees()}. In OUTBREAK, the option appears while changing the duration parameter (option 3). When selected, OUTBREAK will include any sequences of negative indices at the either the beginning or the end of each tree series as a defoliation event, without consideration of either duration or magnitude thresholds. In \texttt{dfoliatR}, the duration threshold is omitted and the magnitude threshold is retained in series-end-events. Each of the 13 events included in these differences did not meet the ``max\_reduction'' parameter (-1.28 NGSI) in \texttt{dfoliatR} and were excluded. In two cases, OUTBREAK included events at the beginning of the series where \texttt{dfoliatR} does not allow truncated events. In four cases, OUTBREAK omitted only the last year of the series because the index was positive. In two cases, \texttt{dfoliatR} omitted possible events because it had already included a positive NGSI excursion after the ``max\_reduction'' year, and since it will only allow one excursion on either side of the max year, the events were omitted due to short duration.
\item
  \emph{Sequential events} (36\%) in which OUTBREAK omitted events that occur one year prior or one year following an identified event. When this occurs, OUTBREAK selects the event with the greatest negative index year. On two trees, OUTBREAK omitted two of three sequential events. While inspecting these differences, we added an option to \texttt{defoliate\_trees()} that would ``bridge'' between sequential events (that surpass magnitude and duration thresholds)into single, long events. We felt that this was ecologically justified, especially for studies of WSBW, because the insect will occasionally defoliate individual trees for decades and reconstructed chronologies show outbreaks as long as \textbf{xxx (ANN CHECK)}.
\item
  \emph{Undetermined differences} (22\%) occurred in cases where OUTBREAK omitted events without clear cause that \texttt{dfoliatR} correctly identified as defoliations.
\item
  \emph{Rounding differences} (2\%) in the indices either omitted or cut short events on two trees. In both cases the indices were very close to zero, and the difference was less than the precision of the raw data measurement.
\end{itemize}

\textbf{(Bridging paragraph?)}

At the site level, OUTBREAK and \texttt{dfoliatR} produce similar time series of percent trees defoliated (Figure \ref{fig:df-obr-site}), which forms the basis for inferring outbreak occurrence, intensity, and duration. Nearly all of the differences between the programs include either more trees or a longer duration of inferred outbreak by \texttt{dfoliatR}. These arise from the inclusion of events omitted by OUTBREAK. These differences translate to greater numbers of trees in defoliation during certain events, or events recorded by single trees during periods of low sample depth.

This evaluation revealed what we believe are short-comings in how OUTBREAK identifies defoliation events on individual trees. In every one of the 102 cases we inspected, we felt that \texttt{dfoliatR} provided a more biologically and statistically appropriate assessment of defoliation. Therefore, \texttt{dfoliatR} also did a better job inferring OUTBREAK events and associated statistics at the site level.

@@

The \texttt{dfoliatR} library (Guiterman, Lynch, and Axelson 2020) is provided free and open source from the Comprehensive R Archive Network (CRAN; \url{https://cran.r-project.org/}). To install \texttt{dfoliatR} from CRAN use

@@

Once \texttt{dfoliatR} is loaded into an \texttt{R} session (via \texttt{library(dfoliatR)}) users can access two sets of tree-ring data to aid in exploring the functions, graphics, and outputs. Each data set consists of individual host-tree series and a local non-host chronology. The host-tree series were standardized using 50-year splines with a 50\% frequency response, while the non-host ring-width data were standardized using 150-year splines with a 50\% frequency response and then averaged via Tukey's biweight robust mean procedure. Host trees from Demijohn Peak (DMJ; 2902 m asl) in the San Juan Mountains of southern Colorado include Douglas-fir (\emph{Pseudotsuga menziesii}) compared against a local non-host ponderosa pine (\emph{Pinus ponderosa}) chronology (Ryerson, Swetnam, and Lynch 2003). The East Fork site (EFK; 2580 m asl) in the Jemez Mountains of northcentral New Mexico includes Douglas-fir and white fir (\emph{Abies concolor}) host trees and a ponderosa pine non-host chronology (Swetnam and Lynch 1993).

@@

The \texttt{dfoliatR} package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes. It is built on the long-accepted host to nonhost comparison methodology used in the 1980s FORTRAN program OUTBREAK (Swetnam, Thompson, and Sutherland 1985; Swetnam and Lynch 1989, 1993). Key benefits are that \texttt{dfoliatR} provides easier control of suppression thresholds, provides additional output tables and high-quality graphics, defines defoliation events more consistently, and operates in the open-source \texttt{R} environment.

>>>>

@@

\usepackage{amsmath}
\usepackage{float}
\usepackage{booktabs}
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \let\@oddfoot\@empty
 \let\@evenfoot\@oddfoot
}
\makeatother

@@

  \begin{abstract}
  We present a new R package to provide dendroecologists with tools to infer, quantify, analyze, and visualize growth suppression events in tree rings. \texttt{dfoliatR} is based on the OUTBREAK program and builds on existing resources in the \texttt{R} and \texttt{dplR} computing environments. It is designed to aid research in the ecology of insect defoliation events and to reconstruct defoliator outbreak chronologies, but can be applied to other studies where host--non-host comparisons are useful. \texttt{dfoliatR} performs an indexing procedure to remove climatic signals in the host-tree series that are represented in the non-host chronologies, or other annually-resolved climate series. It then infers defoliation events in individual trees based on user-specified thresholds. Site-level analyses identify outbreak events that synchronously affect user-defined numbers or proportions of involved host trees. Functions are provided for summary statistics and graphics of tree- and site-level series. We evaluated \texttt{dfoliatR} against OUTBREAK, using eight datasets including 222 host-trees, and found that \texttt{dfoliatR} improves on OUTBREAK with greater user control, identification of defoliation events, computing capacity, and both the statistical summary and graphical outputs. We provide two example data sets and script to enable users to gain familiarity with the package and its capabilities. The source code is available in the Comprehensive R Archive Network (CRAN) and on GitHub.
  \end{abstract}
   \begin{keyword} Dendroecology, spruce budworm, \emph{Choristoneura}, pandora moth, \emph{Coloradia pandora Blake}, larch-bud-moth\end{keyword}
 \end{frontmatter}

@@

Variation in the width and morphology of annual radial growth rings in trees permits dating and quantification of past forest insect defoliator outbreaks. Defoliation can be distinguished from climate- and other disturbance-related influences by comparing ring-width or other annually-resolved features in the wood of host species to that of non-host species or annually-resolved climate records. The effect of defoliation on radial growth of trees has been recognized since the 1860s, and used to reconstruct outbreak regimes since the 1950s (Blais, 1954; Lynch, 2012). It was not until the 1980s, however, that precise dendrochronological techniques were applied for inferring defoliation events and reconstructing defoliator outbreak regimes (Swetnam et al., 1985; Speer, 2010; Lynch, 2012). The first studies (Swetnam et al., 1985; Swetnam and Lynch, 1989, 1993) focused on developing historical outbreak reconstructions of western spruce budworm (WSBW; \emph{Choristoneura freemani} (Lepidoptera: Tortricidae); previously known as \emph{C. occidentalis}). The methodology has since been successfully applied to a wide range of defoliator species, most of which are conifer herbivores, and has evolved in sophistication for a wide range of ecosystem situations (Lynch, 2012).

The main dendrochronological tool for inferring, dating, and characterizing defoliator outbreaks from tree-ring records has been the software routine OUTBREAK (Swetnam et al., 1985; Holmes and Swetnam, 1986; Swetnam and Lynch, 1989). OUTBREAK computes indices (described later in detail) of suppressed growth by subtracting a detrended and standardized climate series (a ``control'' chronology) from individual host-tree detrended and standardized radial growth series after the host and non-host series have been brought to a common variance. The non-host chronology usually consists of a site chronology developed from non-host tree species growing on a climate-sensitive site, but a gridded climate data point series, like the North American Drought Atlas (Cook and Krusic, 2004) also suffices. If the host and non-host species respond similarly to climate (which can and should be tested), the derived series retains variability that the host and non-host series do not have in common, generally the insect signal and some unexplained variability (noise). The user defines a rule base specifying the magnitude and duration that a period of indexed growth suppression must meet or surpass for a period of suppressed growth to be inferred as a defoliation event at the tree level.

Though powerful, OUTBREAK is outdated and increasingly difficult to use in modern computing environments. It was written in FORTRAN V with inherently severe restrictions, as RAM and disk space were limited at that time (256 kb and 10 MB, respectively) and FORTRAN conventions imposed very strict formatting, file naming, and output conventions. The program lacks a graphical interface or capabilities, forcing users to import generated text files into spreadsheets or other software to assess results and perform analyses. Furthermore, OUTBREAK can only handle one test at a time, creating barriers to batch operation and a large burden for researchers with datasets including multiple sites. We developed \texttt{dfoliatR} (Guiterman et al., 2020) as an \texttt{R}- and \texttt{dplR}-based library to overcome these issues.

\texttt{dfoliatR} adds to a growing suite of dendrochronology packages in the \texttt{R} computing environment (R Core Team, 2019). Stemming from the \texttt{dplR} library (Bunn, 2008) that enables \texttt{R} users to read and write an array of tree-ring data formats, standardize ring width series, build and evaluate chronologies, and perform quality control (to name a few), one can now also measure ring widths from scanned images of prepared samples (Lara et al., 2015; Shi et al., 2019), conduct and check crossdating (Bunn, 2010), analyze sub-annual anotomical features (Rathgeber et al., 2011; Campelo et al., 2016), and perform many analytical tests (Zang and Biondi, 2015; Jevšenak and Levanič, 2018). Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events (\texttt{TRADER}: Altman et al., 2014), metrics of growth resilience (\texttt{pointRes}: van der Maaten-Theunissen et al., 2015), and fire history (\texttt{burnr}: Malevich et al., 2018). The key objective of \texttt{dfoliatR} is to provide tools to identify and analyze insect defoliation and outbreak events by building on the methods employed by OUTBREAK. It capitalizes on the robust software already available in \texttt{R} by using \texttt{dplR} data formats for incoming tree-ring series and providing output data formats embodied by the \texttt{tidyverse} (Wickham et al., 2019) that include efficient data manipulation (\texttt{dplyr}: Wickham et al., 2020) and graphics (\texttt{ggplot2}: Wickham, 2016).

In this paper, we describe the statistical methods employed by \texttt{dfoliatR}, compare results to those produced by OUTBREAK, and present an example analysis inlcuding test data sets and script. Users need not have much experience in \texttt{R} to replicate the analyses and graphics as presented. The \texttt{R} code below is executable in an \texttt{R} session once the required libraries are installed and loaded. Support documentation in addition to this paper is provided within the package via standard help menus and on the package website (\url{https://chguiterman.github.io/dfoliatR/}), which includes up-to-date vignettes that describe various routines. Code to generate a preprint of this manuscript, including the \texttt{R} scripts and tabular and graphical output is available from \url{https://github.com/chguiterman/dfoliatR_paper}.

@@

Users can develop these data sets in the software of their choosing, such as \texttt{dplR} or ARSTAN (Cook and Holmes, 1996). It is important that the host-tree data include only one tree-ring series per tree. Both \texttt{dplR} and ARSTAN have options for averaging multiple sample series into a tree-level series.

@@

\begin{figure}
\centering
\includegraphics{D:/Dropbox/dfoliatR_paper/paper_elsevier/Output/tree-plot-default.pdf}
\caption{\label{fig:fig-defol}Default \texttt{dfoliatR} graphics for individual trees, produced by the \texttt{plot\_defol()} function. Dotted lines represent the series length for each tree and colored segments show periods of defoliation. Users can define the cut-off values to determine severe--moderate--minor defoliation intensities, and use standard \texttt{ggplot2} graphics parameters to adjust styles and themes.}
\end{figure}

\begin{figure}
\centering
\includegraphics{D:/Dropbox/dfoliatR_paper/paper_elsevier/Output/site-plot.pdf}
\caption{\label{fig:fig-obr}Default graphics showing outbreak events at the DMJ site. Top panel shows the mean GSI through time, with inferred outbreak events filled. The middle panel shows the percent of trees defoliated, and the bottom panel shows the number of trees through time.}
\end{figure}

@@

Negative departures in the normalized GSI (NGSI, or GSI converted to z-scores) that surpass user-specified thresholds in duration and magnitude are defined as \emph{defoliation events}. As in OUTBREAK, the lowest NGSI value in the particucar sequence being assessed must reach the magnitude threshold. The default setting is -1.28 (NGSI is in units of standard deviation), which was previously determined to be representative of WSBW effects (Swetnam and Lynch, 1989) and is commonly used for other species. The year with the lowest value is termed the ``year of maximum departure'' and becomes a central point in time for assessing other thresholds before being included as a defoliation event. If the year of maximum departure is higher than the threshold (i.e.~\(\textrm{NGSI}_{lowest} > -1.28\)), the sequence being assessed is omitted from the event results.

Event duration is assessed by examining sequences of negative NGSI (that exceed the magnitude threshold) before and after the year of maximum departure year. Each defoliation event is allowed one single-year positive excursion on each side of the year of maximum departure. Duration is computed across the entire sequence that may include these two positive excursions. As in OUTBREAK, the user specifies a duration threshold (minimum number of years) for a departure sequence to be inferred as a defoliation event. The default threshold is eight years, as is commonly used in WSBW studies (Swetnam and Lynch, 1989). If the sequence is shorter than the duration threshold, the sequence is omitted from the event results (i.e., both thresholds must be met). Researchers can, and should, adjust the duration and magnitude parameters accordingly and critically evaluate the results, as insect species vary in the length of their outbreaks and the degree to which they can suppress tree growth.

@@

Diverging from OUTBREAK, \texttt{dfoliatR} allows users to extend defoliation events on individual trees by bridging between sequential events (Figure \ref{fig:fig-bridge}). In cases where two defoliation events are separated by a single year, bridging will link them into a single event. This option was added to \texttt{dfoliatR} during the testing phase of development, when we realized that OUTBREAK deliberately omits sequential, or back-to-back events, even when these events surpass both the magnitude and duration thresholds. Instead, OUTBREAK will select the one sequential event with the lowest negative departure year. In every case we assessed (described below) we felt that the OUTBREAK-omitted defoliation events should have been maintained and recorded. Due in large part to reconstructions using OUTBREAK (see papers cited by Lynch, 2012), we now know considerably more about forest defoliator outbreak regimes than we did in the 1980s when OUTBREAK was under development. We think that two or more prolonged events separated by a single year should in some situations be considered a single event. This is particularly relevant to WSBW and spruce budworm (\emph{C. fumiferana}), for which outbreak regimes have been reconstructed for several geographic areas and for which researchers and forest health experts have gained considerable knowledge (Brookes et al., 1982, and many later publications; Schmitt et al., 1984; Sanders et al., 1985). In many cases, the greatest growth suppression often occurs late in the outbreak due to cumulative effects accrued on a tree's resources (Swetnam and Lynch, 1989, 1993). We urge caution in using the bridging option, however, because it may not be appropriate for all studied insects, such as in situations where impacted stands barely recover from one outbreak before another begins, as with pine processionary caterpillars (\emph{Thaumetopoea pityocampa} (Lepidoptera: Thaumetopoeidae)) (Carus, 2004, 2009).

\begin{figure}
\centering
\includegraphics{D:/Dropbox/dfoliatR_paper/paper_elsevier/Output/bridging.pdf}
\caption{\label{fig:fig-bridge}Examples of sequential defoliation events (bars), wherein two or three events occur separated by single years. In each case OUTBREAK recorded only one of the events, ommitting those with a higher ``max\_defol'' year, or the year of the lowest NGSI value. \texttt{dfoliatR}, by contrast, recorded all of the events, including two sequential events on DMJ26 and EFK22, and three on FNL04. Events recorded by both programs have a darker color, while those recorded only by \texttt{dfoliatR} are lighter in color. Adding the bridging option in \texttt{dfoliatR} links these events (``bridge\_defol'') into single longer events that may more accurately capture the effects of long-term insect herbivory on individual trees.}
\end{figure}

@@

Defoliation of one or a few trees does not constitute an outbreak. To determine when defoliation becomes an \emph{outbreak event}, \texttt{dfoliatR} composites the individual tree defoliation series into a site-level chronology with the \texttt{outbreak()} function. Users have options to define the number and/or the proportion of trees required for an event to be considered an outbreak. Three parameters control whether a defoliation event constitutes an outbreak: the minimum number of trees available, the minimum number of trees recording defoliation, and the percent of trees recording defoliation. The first allows the researcher to make a judgement call as to the confidence ascribed to reduced sample depth toward the ends of their chronologies, thus compensating for the ``fading record problem'' (Swetnam et al., 1999). The second two parameters adjust the scale of defoliation considered to be an outbreak. Absolute numbers of trees and percentages can be applied separately or in conjunction, following filtering conventions in tree-ring fire history studies (Malevich et al., 2018). We urge users to carefully consider the choice of absolute numbers in situations where the number of trees represented in the series varies with time, or the choice of percentages when sample size is small.

@@

We tested \texttt{dfoliatR} against OUTBREAK by comparing NGSI to OUTBREAK's normalized corrected indices for individual trees and years, defoliation status for individual trees and years, and percentage of trees recording outbreaks at the site level. Our tests used standardized ring-width data from eight sites in British Columbia, Colorado, and New Mexico, and author-provided non-host site chronologies.

We detrended host data for both \texttt{dfoliatR} and OUTBREAK using ARSTAN {[}v6.1{]} with cubic smoothing splines (50\% frequency response on 100-150 year wavelengths depending on the site). In both \texttt{dfoliatR} and OUTBREAK we used event thresholds of -1.28 normalized indices, 8 years duration, and allowed for events at the end of series in seven of eight sites. These seven sites were sampled during ongoing outbreak events (Swetnam and Lynch, 1993). We found it necessary to be consistent in how we detrended and what software we employed (e.g., ARSTAN vs \texttt{dplR}) because subtle differences in standardized ring-width indices generated between the programs transferred into differences between \texttt{dfoliatR} and OUTBREAK. In the end, we chose to only use the standardization output files from ARSTAN, which are easily read into \texttt{R} (and then \texttt{dfoliatR}) using the \texttt{dplR} package.

@@

\begin{itemize}
\item
  \emph{Series-end events} (40\% of the total) in which OUTBREAK included ``truncated outbreaks'' (for seven sites) at the end of each series. In \texttt{dfoliatR}, this option is controlled by the ``series\_end\_events'' parameter to \texttt{defoliate\_trees()}. In OUTBREAK, the option appears while changing the duration parameter (option 3). When selected, OUTBREAK will include any sequences of negative indices at the either the beginning or the end of each tree series as a defoliation event, without consideration of either duration or magnitude thresholds. In \texttt{dfoliatR}, the duration threshold is omitted and the magnitude threshold is retained in series-end-events. Each of the 13 events included in these differences did not meet the ``max\_reduction'' parameter (-1.28 NGSI) in \texttt{dfoliatR} and were excluded. In two cases, OUTBREAK included events at the beginning of the series where \texttt{dfoliatR} does not allow truncated events. In four cases, OUTBREAK omitted only the last year of the series because the index was positive, but \texttt{dfoliatR} allowed this single positive excursion. Finally, there were two cases in which \texttt{dfoliatR} omitted possible events because it had already included a positive NGSI excursion after the ``max\_reduction'' year, and since it will only allow one excursion on either side of the max year, the events were omitted due to short duration.
\item
  \emph{Sequential events} (36\%) in which OUTBREAK omitted back-to-back events that occur one year prior or one year following an identified event. When this occurs, OUTBREAK selects the one event sequence with the lowest negative index year (e.g., Figure \ref{fig:fig-bridge}). On two trees, OUTBREAK omitted two of three sequential events. While inspecting these differences, we added an option to \texttt{defoliate\_trees()} that would ``bridge'' between sequential events (that surpass magnitude and duration thresholds) into single, long events. We felt that this was ecologically justified, especially for studies of WSBW, because the insect will occasionally defoliate individual trees for decades and reconstructed chronologies have show outbreaks as long as \textbf{xxx (ANN CHECK)}.
\item
  \emph{Undetermined differences} (22\%) occurred in cases where OUTBREAK omitted events without clear cause that \texttt{dfoliatR} correctly identified as defoliations.
\item
  \emph{Rounding differences} (2\%) in the indices either omitted or cut short events on two trees. In both cases the indices were very close to zero, and the difference was less than the precision of the raw data measurement.
\end{itemize}

At the site level, OUTBREAK and \texttt{dfoliatR} produce similar time series of percent trees defoliated (Figure \ref{fig:fig-df-obr}), which forms the basis for inferring outbreak occurrence, intensity, and duration. In nearly site comparison, \texttt{dfoliatR} included either more events or longer durations of inferred outbreak. These differences arise from the inclusion of tree-level events by \texttt{dfoliatR} that were omitted by OUTBREAK. The differences translate to greater numbers of trees in defoliation during certain events, or events recorded by single trees during periods of low sample depth.

\begin{figure}
\centering
\includegraphics{D:/Dropbox/dfoliatR_paper/paper_elsevier/Output/plot-sites-df-obr.pdf}
\caption{\label{fig:fig-df-obr}Comparison of reconstructed western spruce budworm outbreaks computed by \texttt{dfoliatR} and OUTBREAK. Input parameters were identical between programs. Difference arrise because \texttt{dfoliatR} will identify and record more defoliation events on idnividual trees, and may more accurately capture herbivory dynamics.}
\end{figure}

This evaluation revealed what we believe are short-comings in how OUTBREAK identifies defoliation events on individual trees. In every one of the 102 cases we inspected, we felt that \texttt{dfoliatR} provided a more biologically and statistically appropriate assessment of defoliation. Therefore, \texttt{dfoliatR} also did a better job inferring outbreak events and associated statistics at the site level.

@@

The \texttt{dfoliatR} library is provided free and open source from the Comprehensive R Archive Network (CRAN; \url{https://cran.r-project.org/}). To install \texttt{dfoliatR} from CRAN use

@@

Once \texttt{dfoliatR} is loaded into an \texttt{R} session (via \texttt{library(dfoliatR)}) users can access two sets of tree-ring data to aid in exploring the functions, graphics, and outputs. Each data set consists of individual host-tree series and a local non-host chronology. The host-tree series were standardized using 128-year splines with a 50\% frequency response, while the non-host ring-width data were standardized using 150-year splines with a 50\% frequency response and then averaged via Tukey's biweight robust mean procedure. Host trees from Demijohn Peak (DMJ; 2902 m asl) in the San Juan Mountains of southern Colorado include Douglas-fir (\emph{Pseudotsuga menziesii}) compared against a local non-host ponderosa pine (\emph{Pinus ponderosa}) chronology (Ryerson et al., 2003). The East Fork site (EFK; 2580 m asl) in the Jemez Mountains of northcentral New Mexico includes Douglas-fir and white fir (\emph{Abies concolor}) host trees and a ponderosa pine non-host chronology (Swetnam and Lynch, 1993).

@@

The \texttt{dfoliatR} package provides dendroecologists with tools to infer, quantify, analyze, and visualize tree-ring growth suppression events and to reconstruct forest insect defoliator outbreak regimes. It is built on the long-accepted host to nonhost comparison methodology used in the 1980s FORTRAN program OUTBREAK (Swetnam et al., 1985; Swetnam and Lynch, 1989, 1993). Our evaluation of the two programs revelaed that \texttt{dfolaitR} excelled in identifying defoliation events on single trees, providing researchers with more consistent and biologically-justifiable results. Other key benefits are that \texttt{dfoliatR} provides easier control of suppression thresholds, additional output tables, high-quality and customizable graphics, and operates in the open-source \texttt{R} environment that is stable across computing platforms and under active development and maintence by a large and growing community.
